Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-25 paper.tex
Up paper and link

@@

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{threeparttable}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
My abstract. \newline \textbf{Keywords}: meritocracy, social inequality,
inequality justification, COVID-19
\end{abstract}

@@

@@

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}

@@

Table~\ref{tbl-modelos} presents the results of the multilevel models
estimated for market justice preferences, examining both individuals
(within) and group-level (between) effects. The intraclass correlation
(ICC,
\citeproc{ref-hoxMultilevelAnalysisTechniques2017}{\textbf{hoxMultilevelAnalysisTechniques2017?}})
from the empty model (see Supplementary Material), which decomposes the
variance of market justice preferences, is 0.23, indicating that
approximately 23\% of the variation is attributable to differences
between individuals. Complementary, 77\% of the variation corresponds to
within-individual differences over time.

@@

[H]

@@

\scalebox{1}{
\begin{threeparttable}

@@

\toprule

@@

\midrule

@@

AIC                                  & $20273.132$    & $20260.158$    & $20251.351$    & $20140.808$    & $20002.562$    & $20005.214$    & $19957.283$    & $19986.833$    \\

@@

Log Likelihood                       & $-10128.566$   & $-10123.079$   & $-10117.676$   & $-10061.404$   & $-9991.281$    & $-9991.607$    & $-9963.641$    & $-9964.417$    \\

@@

\bottomrule

@@

\begin{tablenotes}[flushleft]
\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
\end{tablenotes}
\end{threeparttable}

@@

@@

\begin{verbatim}

\begin{table}[h!]
\scalebox{1}{
\begin{threeparttable}
\begin{tabular}{l c c c c}
\toprule
 & Model 9 & Model 10 & Model 11 & Model 12 \\
\midrule
Intercept                                    & $1.605^{***}$ & $1.679^{***}$ & $1.425^{***}$ & $1.444^{***}$ \\
                                             & $(0.123)$     & $(0.112)$     & $(0.117)$     & $(0.117)$     \\
Perception inequality (WE) x Wave            & $0.002$       &               &               &               \\
                                             & $(0.004)$     &               &               &               \\
Justification inequality (WE) x Wave         &               & $0.013^{***}$ &               &               \\
                                             &               & $(0.003)$     &               &               \\
Merit: Effort (WE) x Wave                    &               &               & $-0.012^{**}$ &               \\
                                             &               &               & $(0.004)$     &               \\
Merit: Talent (WE) x Wave                    &               &               &               & $-0.010^{*}$  \\
                                             &               &               &               & $(0.004)$     \\
\midrule
Controls                                     & Yes           & Yes           & Yes           & Yes           \\
AIC                                          & $19995.908$   & $19923.216$   & $19940.577$   & $19932.452$   \\
BIC                                          & $20228.331$   & $20155.638$   & $20172.999$   & $20164.875$   \\
Log Likelihood                               & $-9964.954$   & $-9928.608$   & $-9937.289$   & $-9933.226$   \\
Num. obs.                                    & $8460$        & $8460$        & $8460$        & $8460$        \\
Num. groups: idencuesta                      & $1681$        & $1681$        & $1681$        & $1681$        \\
Var: idencuesta (Intercept)                  & $0.305$       & $0.198$       & $0.137$       & $0.123$       \\
Var: idencuesta perc\_inequality             & $0.004$       & $$            & $$            & $$            \\
Var: idencuesta ola\_num                     & $0.007$       & $0.006$       & $0.006$       & $0.006$       \\
Cov: idencuesta (Intercept) perc\_inequality & $-0.026$      & $$            & $$            & $$            \\
Cov: idencuesta (Intercept) ola\_num         & $-0.032$      & $-0.027$      & $-0.002$      & $-0.000$      \\
Cov: idencuesta perc\_inequality ola\_num    & $0.004$       & $$            & $$            & $$            \\
Var: Residual                                & $0.485$       & $0.484$       & $0.472$       & $0.471$       \\
Var: idencuesta just\_inequality             & $$            & $0.010$       & $$            & $$            \\
Cov: idencuesta (Intercept) just\_inequality & $$            & $-0.022$      & $$            & $$            \\
Cov: idencuesta just\_inequality ola\_num    & $$            & $0.007$       & $$            & $$            \\
Var: idencuesta merit\_effort                & $$            & $$            & $0.022$       & $$            \\
Cov: idencuesta (Intercept) merit\_effort    & $$            & $$            & $-0.024$      & $$            \\
Cov: idencuesta merit\_effort ola\_num       & $$            & $$            & $-0.005$      & $$            \\
Var: idencuesta merit\_talent                & $$            & $$            & $$            & $0.021$       \\
Cov: idencuesta (Intercept) merit\_talent    & $$            & $$            & $$            & $-0.023$      \\
Cov: idencuesta merit\_talent ola\_num       & $$            & $$            & $$            & $-0.005$      \\
\bottomrule
\begin{tablenotes}[flushleft]
\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
\end{tablenotes}
\end{threeparttable}
\end{table}
\end{verbatim}

@@

>>>>

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

@@

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
This study examines the interplay between perceptions of economic
inequality and meritocracy in shaping preferences for market justice in
Chile from 2016 to 2023. Economic inequality has intensified globally,
raising concerns about its social impacts and the legitimacy of
market-based distributions of resources. By utilizing longitudinal data
from the Chilean Longitudinal Social Survey (ELSOC), the research
investigates how subjective assessments of inequality influence
attitudes towards social justice, focusing on the extent to which
individuals believe access to essential services like healthcare,
education, and pensions should depend on personal merit. The findings
suggest that as perceptions of economic inequality rise, individuals
tend to favor redistributive policies, although those with strong
meritocratic beliefs are more likely to justify existing inequalities.
The analysis also highlights the effect of major social movements in
Chile during this period, which have shifted public discourse and
potentially altered perceptions of inequality and meritocracy. This
paper contributes to the understanding of how individual beliefs about
fairness and equity evolve within a socio-political context
characterized by high inequality and extensive free-market principles.
\newline \textbf{Keywords}: Economic inequality, meritocracy, market
justice, Chile, public preferences
\end{abstract}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\#| include: false}
\CommentTok{\#| echo: false}

\ControlFlowTok{if}\NormalTok{ (}\SpecialCharTok{!} \FunctionTok{require}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)) }\FunctionTok{install.packages}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)}

\NormalTok{pacman}\SpecialCharTok{::}\FunctionTok{p\_load}\NormalTok{(tidyverse, }
\NormalTok{               sjmisc, }
\NormalTok{               sjPlot, }
\NormalTok{               lme4, }
\NormalTok{               here, }
\NormalTok{               performance,}
\NormalTok{               influence.ME, }
\NormalTok{               marginaleffects,}
\NormalTok{               MLMusingR,}
\NormalTok{               texreg, }
\NormalTok{               ggdist,}
\NormalTok{               misty,}
\NormalTok{               kableExtra,}
\NormalTok{               ggalluvial, }
\NormalTok{               shadowtext,}
\NormalTok{               MetBrewer,}
\NormalTok{               patchwork,}
\NormalTok{               sjlabelled,}
\NormalTok{               summarytools)}

\FunctionTok{options}\NormalTok{(}\AttributeTok{scipen=}\DecValTok{999}\NormalTok{)}
\FunctionTok{rm}\NormalTok{(}\AttributeTok{list =} \FunctionTok{ls}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

@@

\includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}

@@

Table~\ref{tbl-modelos} presents the results of the multilevel models
estimated for market justice preferences, examining both individuals
(within) and group-level (between) effects. The intraclass correlation
(\citeproc{ref-hoxMultilevelAnalysisTechniques2017a}{Hox et al., 2017})
from the empty model (see Supplementary Material), which decomposes the
variance of market justice preferences, is 0.23, indicating that
approximately 23\% of the variation is attributable to differences
between individuals. Complementary, 77\% of the variation corresponds to
within-individual differences over time.

@@

@@

\scalebox{0.75}{

@@

\hline

@@

\hline

@@

@@

@@

\hline
\multicolumn{9}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}

@@

@@

Table~\ref{tbl-interactions} displays

@@

\begin{center}
\scalebox{0.95}{
\begin{tabular}{l c c c c}
\hline
 & Model 9 & Model 10 & Model 11 & Model 12 \\
\hline
Intercept                                    & $1.605^{***}$ & $1.679^{***}$ & $1.425^{***}$ & $1.444^{***}$ \\
                                             & $(0.123)$     & $(0.112)$     & $(0.117)$     & $(0.117)$     \\
Perception inequality (WE) x Wave            & $0.002$       &               &               &               \\
                                             & $(0.004)$     &               &               &               \\
Justification inequality (WE) x Wave         &               & $0.013^{***}$ &               &               \\
                                             &               & $(0.003)$     &               &               \\
Merit: Effort (WE) x Wave                    &               &               & $-0.012^{**}$ &               \\
                                             &               &               & $(0.004)$     &               \\
Merit: Talent (WE) x Wave                    &               &               &               & $-0.010^{*}$  \\
                                             &               &               &               & $(0.004)$     \\
\hline
Controls                                     & Yes           & Yes           & Yes           & Yes           \\
BIC                                          & $20228.331$   & $20155.638$   & $20172.999$   & $20164.875$   \\
Num. obs.                                    & $8460$        & $8460$        & $8460$        & $8460$        \\
Num. groups: idencuesta                      & $1681$        & $1681$        & $1681$        & $1681$        \\
Var: idencuesta (Intercept)                  & $0.305$       & $0.198$       & $0.137$       & $0.123$       \\
Var: idencuesta perc\_inequality             & $0.004$       & $$            & $$            & $$            \\
Var: idencuesta ola\_num                     & $0.007$       & $0.006$       & $0.006$       & $0.006$       \\
Cov: idencuesta (Intercept) perc\_inequality & $-0.026$      & $$            & $$            & $$            \\
Cov: idencuesta (Intercept) ola\_num         & $-0.032$      & $-0.027$      & $-0.002$      & $-0.000$      \\
Cov: idencuesta perc\_inequality ola\_num    & $0.004$       & $$            & $$            & $$            \\
Var: Residual                                & $0.485$       & $0.484$       & $0.472$       & $0.471$       \\
Var: idencuesta just\_inequality             & $$            & $0.010$       & $$            & $$            \\
Cov: idencuesta (Intercept) just\_inequality & $$            & $-0.022$      & $$            & $$            \\
Cov: idencuesta just\_inequality ola\_num    & $$            & $0.007$       & $$            & $$            \\
Var: idencuesta merit\_effort                & $$            & $$            & $0.022$       & $$            \\
Cov: idencuesta (Intercept) merit\_effort    & $$            & $$            & $-0.024$      & $$            \\
Cov: idencuesta merit\_effort ola\_num       & $$            & $$            & $-0.005$      & $$            \\
Var: idencuesta merit\_talent                & $$            & $$            & $$            & $0.021$       \\
Cov: idencuesta (Intercept) merit\_talent    & $$            & $$            & $$            & $-0.023$      \\
Cov: idencuesta merit\_talent ola\_num       & $$            & $$            & $$            & $-0.005$      \\
\hline
\multicolumn{5}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
\end{tabular}
}
\caption{}
\label{table:coefficients}
\end{center}

@@

\bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
{Analysis}: {Techniques} and {Applications}}.
